They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-15 paper/paper.Rmd
start momentum section

## Distribution of rounds played

@@

```{=tex}
\begin{equation}\protect\hypertarget{eq-series-length}{}{
\chi^2 = \sum^{11}_{i=6} \frac{(P_O(i) - P_E(i))^2}{P_E(i)}, i \in R = [6, 7, 8, 9, 10, 11].
}\label{eq:chi-squ-test}\end{equation}
```
as 16.0 (p-value of 0.0068), I can comfortably reject the constant probability null hypothesis for $p_0 = 0.5$, event at a confidence level of $\alpha = 0.01$.

@@

>>>>

## Distribution of rounds played {#sec:analysis-1}

@@

```{=tex}
\begin{equation}\protect\hypertarget{eq:series-length}{}{
\chi^2 = \sum^{11}_{i=6} \frac{(P_O(i) - P_E(i))^2}{P_E(i)}, i \in R = [6, 7, 8, 9, 10, 11].
}\label{eq:chi-squ-test}\end{equation}
```
as 16.0 (p-value of 0.0068), I can comfortably reject the constant probability null hypothesis for $p_0 = 0.5$, event at a confidence level of $\alpha = 0.01$.

@@

## Momentum {#sec:analysis-2}

```{r}
#| label: cod_round_win_prop_after_x_wins
## http://keyonvafa.com/hot-hand/
get_post_streak_prob <- function(n, k, p = 0.5) {
  tosses <- rbinom(n, 1, p)
  runs <- rle(tosses)
  n_neg_after <- length(which(runs$values == 1 & runs$lengths >= k))
  n_pos_after <- sum(runs$lengths[which(runs$values == 1 & runs$lengths >= k)] - k)

  ## edge case
  if (n %in% cumsum(runs$lengths)[which(runs$values == 1 & runs$lengths >= k)]) {
    n_neg_after <- n_neg_after - 1
  }

  n_pos_after / (n_pos_after + n_neg_after)
}

simulate_post_streak_prob <- function(sims = 10000, seed = 42, ...) {
  withr::local_seed(seed)
  rerun(
    sims,
    get_post_streak_prob(...)
  ) |> 
    flatten_dbl() |> 
    mean(na.rm = TRUE)
}

cod_round_streaks <- cod_rounds |> 
  group_by(series_id, team) |> 
  mutate(
    won_prior_round1 = lag(win_round, n = 1, default = NA),
    won_prior_round2 = lag(win_round, n = 2, default = NA),
    won_prior_round3 = lag(win_round, n = 3, default = NA),
    won_prior_round4 = lag(win_round, n = 4, default = NA),
    won_prior_round5 = lag(win_round, n = 5, default = NA)
  ) |> 
  ungroup() |>
  select(
    series_id,
    team,
    is_offense,
    round,
    n_rounds,
    win_series,
    win_round,
    starts_with('won_prior_round')
  )

postprocess_round_streaks <- function(k, probs_nx_k, ...) {

  cols <- sprintf('won_prior_round%d', 2:k)
  col_syms <- syms(cols)
  cod_round_streaks_after_x <- cod_round_streaks |> 
    drop_na(!!!col_syms) |> 
    count(n_rounds, ..., win_series, win_round, won_prior_round1, !!!col_syms, sort = TRUE)

  cod_round_win_prop_after_z <- cod_round_streaks_after_x |> 
    filter(won_prior_round1, !!!col_syms)

  suppressMessages(
    cod_round_win_prop_after_z_with_probs <- cod_round_win_prop_after_z |> 
      group_by(win_series, n_rounds) |> 
      mutate(
        total = sum(n),
        prop = n / total
      ) |> 
      ungroup() |> 
      filter(win_round) |> 
      select(n_rounds, ..., win_series, n, total, prop) |> 
      arrange(n_rounds, ...) |> 
      inner_join(
        probs_nx_k
      )
  )

  cod_round_win_prop_after_z_with_probs |> 
    nest(data = -c(n_rounds, win_series, ...)) |> 
    mutate(
      gt_p_value = map_dbl(
        data, 
        ~binom.test(.x$n, .x$total, .x$prior_p, alternative = 'greater') |> 
          broom::tidy() |>
          pull(p.value)
      ),
      lt_p_value = map_dbl(
        data, 
        ~binom.test(.x$n, .x$total, .x$prior_p, alternative = 'less') |> 
          broom::tidy() |>
          pull(p.value)
      ),
      gt_p_value_adj = map_dbl(
        data, 
        ~binom.test(.x$n, .x$total, .x$expected_poststreak_p, alternative = 'greater') |> 
          broom::tidy() |>
          pull(p.value)
      ),
      lt_p_value_adj = map_dbl(
        data, 
        ~binom.test(.x$n, .x$total, .x$expected_poststreak_p, alternative = 'less') |> 
          broom::tidy() |>
          pull(p.value)
      ),
      gt_is_signif = gt_p_value <= 0.05,
      lt_is_signif = lt_p_value <= 0.05,
      gt_is_signif_adj = gt_p_value_adj <= 0.05,
      lt_is_signif_adj = lt_p_value_adj <= 0.05
    ) |> 
    unnest(data)
}

summarize_cod_win_prop_after_x_wins <- function(k) {
  probs_nx_k <- crossing(
    n_rounds = 7L:11L,
    win_series = c(TRUE, FALSE)
  ) |> 
    mutate(
      is_valid = case_when(
        win_series ~ TRUE,
        (n_rounds - !!k - 6L) > 0L ~ TRUE,
        TRUE ~ FALSE
      )
    ) |> 
    filter(is_valid) |> 
    select(-is_valid) |> 
    mutate(
      prior_p = (ifelse(win_series, 6L, n_rounds - 6L) - k) / (n_rounds - k),
      expected_poststreak_p = map2_dbl(
        n_rounds, prior_p, 
        ~simulate_post_streak_prob(n = ..1, k = !!k, p = ..2)
      )
    )

  postprocess_round_streaks(
    k = k,
    probs_nx_k = probs_nx_k
  )
}

cod_round_win_prop_after_x_wins <- tibble(win_streak = 2L:4L) |>
  mutate(
    data = map(win_streak, summarize_cod_win_prop_after_x_wins)
  ) |>
  unnest(data)
```

```{r}
#| label: cod_round_win_prop_after_x_wins-table
cod_round_win_prop_after_x_wins |> 
  mutate(
    across(
      matches('p_value'), 
      ~case_when(
        .x <= 0.01 ~ '<=0.01', 
        round(.x, 2) == 1 ~ '1',
        TRUE ~ as.character(round(.x, 2))
      )
    )
  ) |>
  # filter(win_streak == 3) |> 
  select(
    win_streak,
    win_series,
    n_rounds,
    total,
    prop,
    prior_p,
    expected_poststreak_p,
    gt_p_value,
    gt_p_value_adj
  ) |> 
  gt::gt() |> 
  gt::fmt_percent(
    decimals = 0,
    columns = c(prop, prior_p, expected_poststreak_p)
  ) |> 
  gt::cols_label(
    win_streak = 'Streak',
    win_series = 'Win Series?',
    n_rounds = 'Series lasts $i$ rounds',
    total = 'N',
    prop = 'Obs.',
    prior_p = 'Naive',
    expected_poststreak_p = 'Discounted',
    gt_p_value = 'Naive',
    gt_p_value_adj = 'Discounted'
  ) |> 
  gt::tab_spanner(
    label = '%s',
    columns = c(prop, prior_p, expected_poststreak_p)
  ) |> 
  gt::tab_spanner(
    label = 'Binom. > p-value',
    columns = c(gt_p_value, gt_p_value_adj)
  )
```

Despite the implication that one can model team round win probability in CoD SnD as a constant, anecdotally many believe in momentum i.e. streaks not due to random chance. I now evaluate win frequency in rounds following a streak of length $k$. I account for the length of the series ($i$ rounds), modeling the expected round win probability after a streak of length $k$ as

```{=tex}
\begin{equation}\protect\hypertarget{eq:pki}{}{
  \hat{P}(\text{win round | k-round winning streak, series lasting i rounds})=\{
  \begin{array}{cl}
    \frac{6 - k}{i - k}, & \text{team wins series} \\
    \frac{11 - 6 - k}{i - k}, & \text{team loses series}
  \end{array}.
}\label{eq:pki}\end{equation}
```
Miller et al. [-@miller2018] provide evidence for negative recency suggesting that

```{=tex}
\begin{equation}\protect\hypertarget{eq:pkid}{}{
  \begin{array}{rcl}
  \hat{P}_{ki,\text{diff}} & = & \hat{P}(\text{win round | k-round winning streak, series lasting i rounds}) -  \\
    & & \hat{P}(\text{lose round | k-round losing streak, series lasting i rounds}) \\
    E[\hat{P}_{ki,\text{diff}}] & \neq & 0
  \end{array}
}\label{eq:pkid}\end{equation}
```
given a deterministic i.i.d. Bernoulli probability $\hat{p}(\text{win round | k-round winning streak, series lasting i rounds)}$ (Equation \ref{eq:pki}.[^13]

[^13]: Note that these probabilites are separate from those in Section \\ref{sec:analysis-1}.

Given that people typically perceive streaks as beginning after the third success (or failure) at minimum [@carlson2007], I focus on round win percentages in rounds immediately following streaks of three wins, $k = 3$.[^14]

[^14]: Three happens to also be a reasonable number for series that lasts at maximum 11 rounds.

```{r}
#| label: cod_round_win_prop_after_x_wins_and_rounds
summarize_cod_win_prop_after_x_wins_y_rounds <- function(k) {
  probs_nx_k <- crossing(
    n_rounds = 7L:11L,
    round = 3L:11L
  ) |> 
    filter(round >= (k + 1L)) |> 
    mutate(
      prior_p = 6L / n_rounds,
      expected_poststreak_p = map2_dbl(
        n_rounds, prior_p, 
        ~simulate_post_streak_prob(n = ..1, k = !!k, p = ..2)
      )
    )

  postprocess_round_streaks(
    k = k,
    probs_nx_k = probs_nx_k,
    round
  )
}

cod_round_win_prop_after_x_wins_y_rounds <- tibble(win_streak = 2L:4L) |>
  mutate(
    data = map(win_streak, summarize_cod_win_prop_after_x_wins_y_rounds)
  ) |>
  unnest(data)
```
